### 2014-04-15 paper/abstract.tex
typographical fun

%!TEX root = series of tubes.tex
\begin{abstract}
3D printers offer extraordinary flexibilty for prototyping the shape and mechanical function of objects.  We investigate how 3D models can be modified to facilitate the creation of \emph{interactive} objects that offer dynamic input and output.
%However, a 3D printer's role in prototyping interactive objects is still not clear.  
We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material from 3D models to form pipes and chambers. We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, path constraints, and inserted media.  We then present \systemnamenospace, a tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.  We use two distinct routing algorithms.  The first allows users to define the pipes' terminals, and then uses path routing and physics-based simulation to minimize pipe bending energy, allowing easy insertion of media post-print.  The second allows users to supply a desired internal shape to which we fit a pipe route: for this we developed a novel graph-based routing algorithm.  We present several prototypes we created using our tool to show its flexibility and potential.
\end{abstract}

>>>>

%!TEX root = series of tubes.tex
\begin{abstract}
3D printers offer extraordinary flexibility for prototyping the shape and mechanical function of objects.  We investigate how 3D models can be modified to facilitate the creation of \emph{interactive} objects that offer dynamic input and output.
%However, a 3D printer's role in prototyping interactive objects is still not clear.  
We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material from 3D models to form pipes and chambers. We describe this new design space of pipes for interaction design, where variables include openings, path constraints, topologies, and inserted media.  We then present \systemnamenospace, a tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.  We use two distinct routing algorithms.  The first allows users to define the pipes' terminals, and then uses path routing and physics-based simulation to minimize pipe bending energy, allowing easy insertion of media post-print.  The second allows users to supply a desired internal shape to which we fit a pipe route: for this we developed a novel graph-based routing algorithm.  We present several prototypes we created using our tool to show its flexibility and potential.
\end{abstract}

### 2014-04-15 paper/designspace.tex
typographical fun

\section{The Design Space of Pipes}
Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general purpose tools currently exist for this purpose. To guide the design of such tool, we now provide a design space of the types of pipes that a maker may want to utilize. We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.  We describe the space (see Figure \ref{fig:pipespace}).

@@

\subsection{Openings}
Pipes inside solid objects can be either \emph{open} to the outside on both ends, \emph{semi-enclosed}, or \emph{fully enclosed} (see Figure \ref{fig:pipespace}). For interactive devices, openings can be either user-facing (in places a user would for example touch a model); or system-facing, where sensors and actuators may be connected.
%\george{difference between open/return not clear}
%Pipes can have four opening styles: open, return, semi-closed, and fully enclosed .  As we are interested in interactive devices, we distinguish a ``user side"---part of the surface or interior of an object facing its user for interaction; and a ``system side"---where other hardware such as electronics may be connected.  We refer to a pipe which is ``open on the user side'' to have its contents \emph{physically accessible} to a user \emph{at interaction time}.  ``Closed on the user side'' implies physically cut off from the user.  Similarly with the system side.  User or system manipulation of pipe contents are possible even without physical access: e.g., via soundwaves. \george{confusing}

@@

\emph{Open} pipes may be used to create electronic circuitry.  For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (see Figure \ref{fig:openings}a).
%
\emph{Semi-enclosed} pipes are capped on one side and could be used to create tactile output. By fabricating a cap from a malleable material and attaching a pump to the open end, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}b).
%
\emph{Fully enclosed} pipes (cavities) are entirely internal to the object.  Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}c.  If printed in parts and assembled post-print, enclosed pipes could also contain water or particles that would otherwise fall out.

\subsection{Path Constraints}
Depending on the purpose of the pipe, its geometry may be constrained to specific locations within the 3D model. An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is irrelevant. This is the case for circuit schematics, where the electronic components have a fixed location, but many routing solutions are valid. In contrast, for a \emph{path-constrained} pipe the shape of the path is important. For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:teaser}, right). A \emph{fully constrained} pipe requires both specific terminals as well as a specific path. If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, she can use a fully-constrained path (Figure \ref{fig:neon}). Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points. A cavity to modify an object's weight is an example of an unconstrained pipe.

\subsection{Pipe Topologies}
Pipe network topologies enable \emph{splitting} or \emph{mixing} of media (see Figure \ref{fig:pipespace}).  \emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different sensing locations for our touch-sensitive toys, where one terminal is used for swept-frequency capacitive sensing (Figure~\ref{fig:examples}a).

\subsection{Media in Pipes}
After printing, different media can be inserted into pipes to change affordances and capabilities. We consider \emph{gas}, \emph{liquids}, \emph{solids}, \emph{particulates}, and \emph{threadables}. 
%
Compressible fluids (\emph{``gas''}) and incompressible fluids (\emph{``liquids''})  inside semi-enclosed pipes can create haptic feedback; pressure can also be sensed to act as an input~\cite{Slyper-shape}. Gases can be used as carriers for scents or fog. Conductive liquids like copper paint, when dried, turn pipes into conductive paths and can also fix electronic components in place.
%
Filling pipes with \emph{solids} at print time enables designers to exploit the difference in material properties between pipe and contents. For example, in Printed Optics~\cite{Willis-printedoptics}, different refractive indices between solid cladding material and transparent material inside pipes allowed light transport along pipes for input and output.
%
\emph{Particulates}, either printed in-place or inserted, can be of varying densities and can contain particles of various and variable size.  A single large particle can be used for display.  Small particles in variable amounts of fluid can provide variable stiffness haptics via jamming \cite{Follmer-jamming}.
%
\emph{Threadable} media, such as regular wire, electroluminescent wire, or commercial fiberoptic cables, can be threaded through pipes post-printing. Threading such media enables addition of conduction, illumination, or optical sensing using materials that can easily be procured, but cannot themselves be printed, e.g., on consumer printers with a single material. 
% This empowers inexpensive printers: for example, a Printed Optics-style interface can be created on a consumer-grade 3D printer with pipes and fiber optic cable.

>>>>

\section{The Design Space of Pipes}
Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general-purpose tools currently exist for this purpose. To guide the design of such a tool, we now provide a design space of the types of pipes that a maker may want to utilize. We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.  We describe the space (see Figure \ref{fig:pipespace}).

@@

\subsection{Openings}
Pipes inside solid objects can be either \emph{open} to the outside on both ends, \emph{semi-closed}, or \emph{fully enclosed} (see Figure \ref{fig:pipespace}). For interactive devices, openings can be either user-facing, where a user would, for example, touch a model; or system-facing, where sensors and actuators may be connected.

@@

\emph{Open} pipes may be used to create electronic circuitry.  For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (see Figure \ref{fig:openings}a).
%
\emph{Semi-closed} pipes are capped on one side and could be used to create tactile output. By fabricating a cap from a malleable material and attaching a pump to the open end, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}b).
%
\emph{Fully enclosed} pipes (cavities) are entirely internal to the object.  Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}c.  If printed in parts and assembled post-print, enclosed pipes could also contain water or particles that would otherwise fall out.

\subsection{Path Constraints}
Depending on the purpose of the pipe, its geometry may be constrained to specific locations within the 3D model. An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is irrelevant. This is the case for circuit schematics, where the electronic components have fixed locations, but many routing solutions are valid. In contrast, for a \emph{path-constrained} pipe the shape of the path is important. For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:teaser}, right). A \emph{fully constrained} pipe requires specific terminals as well as a specific path. If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, she can use a fully-constrained path (Figure \ref{fig:neon}). Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points. A cavity to modify an object's weight is an example of an unconstrained pipe.

\subsection{Pipe Topologies}
Pipe network topologies enable \emph{splitting} or \emph{mixing} of media (see Figure \ref{fig:pipespace}).  \emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different sensing locations for our touch-sensitive toys, where one terminal is used for swept-frequency capacitive sensing (Figure~\ref{fig:teaser}, center).

\subsection{Media in Pipes}
After printing, different media can be inserted into pipes to change affordances and capabilities. We consider \emph{gas}, \emph{liquids}, \emph{solids}, \emph{particulates}, and \emph{threadables}. 
%
Compressible fluids (\emph{``gas''}) and incompressible fluids (\emph{``liquids''})  inside semi-closed pipes can create haptic feedback; pressure can also be sensed to act as an input~\cite{Slyper-shape}. Gases can be used as carriers for scents or fog. Conductive liquids like copper paint, when dried, turn pipes into conductive paths and can also fix electronic components in place.
%
Filling pipes with \emph{solids} at print time enables designers to exploit the difference in material properties between pipe and contents. For example, in Printed Optics~\cite{Willis-printedoptics}, different refractive indices between solid cladding material and transparent material inside pipes allow light transport along pipes for input and output.
%
\emph{Particulates}, either printed in-place or inserted, can be of varying densities and can contain particles of various and variable size.  A single large particle can be used for display.  Small particles in variable amounts of fluid can provide variable stiffness haptics via jamming \cite{Follmer-jamming}.
%
\emph{Threadable} media, such as regular wire, electroluminescent wire, or commercial fiberoptic cables, can be threaded through pipes post-printing. Threading such media enables addition of conduction, illumination, or optical sensing using materials that can easily be procured, but cannot themselves be printed, e.g., on consumer printers with a single material. 
% This empowers inexpensive printers: for example, a Printed Optics-style interface can be created on a consumer-grade 3D printer with pipes and fiber optic cable.

### 2014-04-15 paper/introduction.tex
typographical fun

Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
%and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects. 
Existing approaches include Printed Optics \cite{Willis-printedoptics}, which print light pipes using optically clear material, and Sauron \cite{Savage-sauron}, which uses computer vision to track the movment of components on the inside of a hollow object. These systems exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities.
%is based on computer vision, but they are limited to visible (reflecting light off fingers, mechanical motion of components) input and visual (screen-like) output; they do not give makers freedom to integrate haptic feedback or other inputs or outputs.

@@

We describe a new design space of pipes and hollow chambers for the purpose of interaction design, where variables include openings, path constraints, topologies, and inserted media. By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized. For example, injecting conductive paint can turn pipes into conductors, which can in turn be used to integrate electronic components into prints. Electroluminescent (EL) wire can be threaded through pipes to create computer-controlled visual output. And air can be pumped through pipes to produce haptic effects.  Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.  We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to or through arbitrary locations on a device.

@@

Figure \ref{fig:teaser} shows a collection of interactive objects with various functionalities including haptic feedback, neon sign-like visual output, and selective touch-sensitivity.  All these objects have had the electronic components required to function redirected onto their surfaces using pipes.  But generating the appropriate pipes for these examples is not trivial.  For one, most 3D modelling tools do not provide extensive tools for modelling the  {\em interior} of 3D models. Furthermore, careful consideration needs to go into the design and routing of the pipes. Pipes need maximum possible bend radius to ease the insertion of media post-print.  Independent pipes cannot intersect, as this could lead to electrical shorting or other problems.  Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.  Pipes for applications like illumunated EL wire signs must allow a single pipe to follow a specific path layout (e.g., for spelling out text).

We contribute PipeDream, an interactive design tool that offers 3D point-to-point routing of internal pipes based on a physical simulation of flexible rods. Our system addresses the various challenges that makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles. We then present a set of example objects, designed with the help of our modeling tool, that demonstrate different interaction modalities. We conclude by discussing limitations and outlining areas for future exploration.

>>>>

Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
%and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects. 
Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses light pipes printed in optically clear material, and Sauron \cite{Savage-sauron}, which uses computer vision to track the movement of components on the inside of a hollow object. These systems exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities.
%is based on computer vision, but they are limited to visible (reflecting light off fingers, mechanical motion of components) input and visual (screen-like) output; they do not give makers freedom to integrate haptic feedback or other inputs or outputs.

@@

We describe a new design space of pipes for the purpose of interaction design, where variables include openings, path constraints, topologies, and inserted media. By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized. For example, injecting conductive paint can turn pipes into conductors, which can in turn be used to integrate electronic components into prints. Electroluminescent (EL) wire can be threaded through pipes to create computer-controlled visual output. And air can be pumped through pipes to produce haptic effects.  Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.  We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to or through arbitrary locations on a device.

@@

Figure \ref{fig:teaser} shows a collection of interactive objects with various functionalities including haptic feedback, neon sign-like visual output, and selective touch-sensitivity.  All these objects have had the electronic components required to function redirected using pipes.  But generating the appropriate pipes for these examples is not trivial.  For one, most 3D modelling tools do not provide extensive tools for modelling the  {\em interior} of 3D models. Furthermore, careful consideration needs to go into the design and routing of the pipes. Pipes need maximum possible bend radius to ease the insertion of media post-print.  Independent pipes cannot intersect, as this could lead to electrical shorting or other problems.  Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.  Pipes for applications like illuminated EL wire signs must allow a single pipe to follow a specific path layout (e.g., for spelling out text).

We contribute PipeDream, an interactive design tool that offers 3D point-to-point routing of internal pipes based on physical simulation of flexible rods. Our system addresses the various challenges that makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles. We then present a set of example objects, designed with the help of our modeling tool, that demonstrate different interaction modalities. We conclude by discussing limitations and outlining areas for future exploration.

### 2014-04-15 paper/tubes.tex
typographical fun

% Designing objects which, for example, are touch sensitive in particular areas or contain electronic components, requires precise location and sizing of pipe endpoints.  The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
To create endpoint-constrained pipes, in which the location and shape of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface. The system automatically routes the pipe, and displays a preview.  Using sliders, the user can independently adjust the starting and ending radii of the pipe.  This is useful for, e.g, our breathing bunny (Figure \ref{fig:examples}b), whose pipes must fit a piece of hardware on one end, but on the other end should have as large an interactive surface as possible.

@@

PBD simulations rarely converge to a steady state, so we simply halt after a fixed number of timesteps (250). Our voxel grid currently uses a resolution of 128x128x128 voxels, which enables interactive operations, but this parameter can be adjusted for different model requirements. The simulation takes about a second for most of our examples.

@@

%In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.  This kind of graph supports an Euler tour: a path that travels every edge exactly once, beginning and ending at the two nodes of odd degree (in this case, the user-selected start and end nodes).  We create a graph based on the user-specified geometry data, with edges for all line segments; then add edges to connect graph components and make graph the semi-Eulerian.  We find an Euler tour on the modified graph, and thicken the path to create pipes.  
Our technique for routing this path is inspired by work in generating continuous line illustrations from images~\cite{Wong-continuousline,Bosch-tsp}.  Our routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described by the user's input, much as a postman needs to walk along every road at least once to deliver mail.  In our relaxation, we allow the creation of new edges (i.e., the postman may jump from building to building in addition to using existing roads).
%
%If path components are disconnected, we create edges that connect them; additionally we can create edges connecting odd-degree vertices rather than simply retracing existing edges.  In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).  
%We offer a short algorithm here, with a more mathematically precise definition and associated proof in supplemental materials appendix.

@@

{\bf Euler Tour:} To find an Euler tour, we use Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}}, with a modification to preserve continuity: at a given node, from multiple candidate edges, we prefer the edge(s) with the smallest angular deviation from the incoming edge (i.e., we prefer to pass straight through a node, if possible).  This minimizes turns in the final artifact, which eases support material removal and assembly.  Our EL wire in Figure \ref{fig:UIST} follows an Euler tour through the input nodes.

@@

\begin{figure}[t]
\centering
    \includegraphics[width=3.4in]{figures/cap.pdf}
\caption{The process to ``cap'' the end of a pipe.  In (a), a pipe end is placed on the surface of the mesh.  In (b), we create a sphere whose radius is slightly larger than the radius of the pipe, centred at the tube's endpoint.  In (c), we consider the intersection of the sphere and the pipe, and create an offset surface of the mesh.  Finally, in (d) we intersect with the offset surface.  This is kept as a cap for the pipe in mesh subtraction.}
\label{fig:cap}
\end{figure}

@@

\subsection{Multi-Jetting}
On multi-jetting machines, bridging is not possible and stepover potential is much less, because the machine is laying down droplets of liquid that are later UV cured.  Unsupported overhangs can be approximately $14^{\circ}$ on our Objet printer.  Multi-jetting and higher-end FDM machines use a secondary support material that must be manually removed afterwards. Some support can be melted out (Projet wax material); other support can be dissolved in lye or blasted with high-pressure water. Complex pipe geometries can make water blasting impossible and other processes more labor- and time-intensive.  All models in Figure \ref{fig:teaser}b required support material removal post-print: the neon sign and maze were additionally cut into two pieces to give better access to the support.

\subsection{Powder-based and Stereolithography}
Powder-based and stereolithography (SLA) machines work with a bed of powder or liquid, respectively.  The material at the surface is fused together via glue droplets (for powder) or a UV laser (for SLA).  The remaining powder or liquid serves as support for the model.  These types of printers are perhaps even more suitable for pipe creation than either FDM or multi-jetting printers, as the supporting liquid or loose powder is simpler to remove than structural plastic or UV cured resin.  We hope to experiment with them in the future.

>>>>

% Designing objects which, for example, are touch sensitive in particular areas or contain electronic components, requires precise location and sizing of pipe endpoints.  The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
To create endpoint-constrained pipes, in which the location of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface. The system automatically routes the pipe, and displays a preview.  Using sliders, the user can independently adjust the starting and ending radii of the pipe.  This is useful for, e.g, our breathing bunny (Figure \ref{fig:examples}b), whose pipes must fit a piece of hardware on one end, but on the other end should have as large an interactive surface as possible.

@@

PBD simulations rarely converge to a steady state, so we simply halt after a fixed number of timesteps (250). Our voxel grid currently uses a resolution of 128x128x128 voxels, which enables interactive operations, but this parameter can be adjusted for different model requirements. The simulation takes about a second for our examples.

@@

%In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.  This kind of graph supports an Euler tour: a path that travels every edge exactly once, beginning and ending at the two nodes of odd degree (in this case, the user-selected start and end nodes).  We create a graph based on the user-specified geometry data, with edges for all line segments; then add edges to connect graph components and make graph the semi-Eulerian.  We find an Euler tour on the modified graph, and thicken the path to create pipes.  
Our technique for routing this path is inspired by work in generating continuous line illustrations from images~\cite{Bosch-tsp,Wong-continuousline}.  Our routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described by the user's input, much as a postman needs to walk along every road at least once to deliver mail.  In our relaxation, we allow the creation of new edges (i.e., the postman may jump from building to building in addition to using existing roads).
%
%If path components are disconnected, we create edges that connect them; additionally we can create edges connecting odd-degree vertices rather than simply retracing existing edges.  In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).  
%We offer a short algorithm here, with a more mathematically precise definition and associated proof in supplemental materials appendix.

@@

{\bf Euler Tour:} To find an Euler tour, we use Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} with a modification to preserve continuity: at a given node, from multiple candidate edges, we prefer the edge(s) with the smallest angular deviation from the incoming edge (i.e., we prefer to pass straight through a node, if possible).  This minimizes turns in the final artifact, which eases support material removal and assembly.  Our EL wire in Figure \ref{fig:UIST} follows an Euler tour through the input nodes.

@@

\begin{figure}[t]
\centering
    \includegraphics[width=3.4in]{figures/cap.pdf}
\caption{The process to ``cap'' the end of a pipe.  In (a), a pipe end is placed on the surface of the mesh.  In (b), we create a sphere whose radius is slightly larger than the radius of the pipe, centred at the pipe's endpoint.  In (c), we consider the intersection of the sphere and the pipe, and create an offset surface of the mesh.  Finally, in (d) we intersect with the offset surface.  This is kept as a cap for the pipe in mesh subtraction.}
\label{fig:cap}
\end{figure}

@@

\subsection{Multi-Jetting}
On multi-jetting machines, bridging is not possible and stepover potential is much less, because the machine is laying down droplets of liquid that are later UV cured.  Unsupported overhangs can be approximately $14^{\circ}$ on our Objet printer.  Multi-jetting and higher-end FDM machines use a secondary support material that must be manually removed afterwards. Some support can be melted out (Projet wax material); other support can be dissolved in lye or blasted with high-pressure water. Complex pipe geometries can make water blasting impossible and other processes more labor- and time-intensive.  All black and clear models in Figure \ref{fig:teaser} required support removal post-print: the neon sign and maze were additionally cut into two pieces to give better access to the support.

\subsection{Powder-based and Stereolithography}
Powder-based and stereolithography (SLA) machines work with a bed of powder or liquid, respectively.  The material at the surface is fused together via glue droplets (for powder) or a UV laser (for SLA).  The remaining powder or liquid serves as support for the model.  These types of printers are perhaps even more suitable for pipe creation than either FDM or multi-jetting printers, as the supporting liquid or loose powder is simpler to remove than structural plastic or UV cured material.  We hope to experiment with them in the future.
